---
title: "RNA-seq Differential Expression Report"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: true
    toc_depth: 2
    mathjax: null
params:
  outdir: "results"
  cfg: NULL
  forced: false
  run_timestamp: ""
  batch_vars: ""
  contrast: ""
---

```{r setup, include=FALSE}
# WHY this chunk: central config for every figure/table, artifact paths, and
# graceful capability detection (plotly/DT power the interactive views; static
# PNG/kable are the fallbacks so the report renders even without them).
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
have_plotly <- requireNamespace("plotly", quietly = TRUE)
have_DT     <- requireNamespace("DT", quietly = TRUE)

outdir    <- params$outdir
cfg       <- params$cfg
p_qc      <- file.path(outdir, "qc", "qc_results.tsv")
p_varmeta <- file.path(outdir, "qc", "variance_vs_metadata.tsv")
p_pca_tsv <- file.path(outdir, "tables", "pca_scores.tsv")
p_pca_pv  <- file.path(outdir, "tables", "pca_percentvar.txt")
p_pca     <- file.path(outdir, "plots", "pca.png")
p_pca_bc  <- file.path(outdir, "plots", "pca_batch_corrected.png")
p_vm_png  <- file.path(outdir, "plots", "variance_vs_metadata.png")
p_volcano <- file.path(outdir, "plots", "volcano.png")
p_de      <- file.path(outdir, "tables", "de_results.tsv")
p_ora     <- file.path(outdir, "tables", "enrichment_ora.tsv")
p_gsea    <- file.path(outdir, "tables", "enrichment_gsea.tsv")
p_dot     <- file.path(outdir, "plots", "enrichment_dotplot.png")

read_tsv_safe  <- function(p, ...) if (file.exists(p)) utils::read.delim(p, check.names = FALSE, ...) else NULL
img_tag <- function(p, alt = "") if (file.exists(p)) sprintf('<img src="%s" alt="%s" style="max-width:100%%;border-radius:8px;">\n', normalizePath(p), alt) else ""

PAL <- list(up = "#e05252", down = "#3f7fbf", ns = "#c9ccd1",
            pass = "#2e9e5b", warn = "#e0a800", fail = "#d64545")
```

```{css, echo=FALSE}
/* ---- Design system: clean, modern, light/dark aware ---------------------- */
:root{
  --bg:#f6f8fb; --card:#ffffff; --ink:#1f2733; --muted:#5b6675; --line:#e4e8ef;
  --accent:#4f46e5; --accent2:#0ea5e9;
  --pass:#2e9e5b; --pass-bg:#e7f6ec; --warn:#e0a800; --warn-bg:#fdf6e3; --fail:#d64545; --fail-bg:#fdeaea;
}
@media (prefers-color-scheme: dark){
  :root{ --bg:#0f141b; --card:#161d27; --ink:#e6eaf0; --muted:#9aa6b6; --line:#26303c;
         --pass-bg:#12301f; --warn-bg:#302811; --fail-bg:#301616; }
}
body{ background:var(--bg); color:var(--ink);
  font-family:-apple-system,BlinkMacSystemFont,"Segoe UI",Roboto,Helvetica,Arial,sans-serif;
  line-height:1.55; }
.main-container{ max-width:1080px; padding:0 22px 80px; margin:0 auto; }
h1.title, .title{ display:none; } /* replaced by custom hero */
/* table of contents styled as a clean card */
#TOC{ background:var(--card); border:1px solid var(--line); border-radius:12px;
  padding:14px 18px; margin:18px 0; box-shadow:0 1px 2px rgba(0,0,0,.04); }
#TOC ul{ list-style:none; padding-left:14px; margin:4px 0; }
#TOC > ul{ padding-left:0; }
#TOC li a{ text-decoration:none; color:var(--muted); }
#TOC li a:hover{ color:var(--accent); }
h1,h2,h3{ color:var(--ink); font-weight:680; letter-spacing:-0.01em; }
h2{ margin-top:2.4em; padding-bottom:.3em; border-bottom:2px solid var(--line); }
a{ color:var(--accent); }
/* hero banner */
.hero{ background:linear-gradient(120deg,var(--accent),var(--accent2)); color:#fff;
  border-radius:16px; padding:26px 30px; margin:22px 0 8px; box-shadow:0 10px 30px rgba(79,70,229,.25); }
.hero h1{ color:#fff; margin:0 0 6px; font-size:30px; }
.hero .meta{ opacity:.95; font-size:14px; display:flex; flex-wrap:wrap; gap:18px; }
.hero .meta b{ font-weight:700; }
/* cards + stat tiles */
.card{ background:var(--card); border:1px solid var(--line); border-radius:12px;
  padding:16px 18px; box-shadow:0 1px 2px rgba(0,0,0,.04); }
.tiles{ display:flex; flex-wrap:wrap; gap:14px; margin:14px 0; }
.tile{ flex:1 1 150px; background:var(--card); border:1px solid var(--line);
  border-radius:12px; padding:16px 18px; text-align:center; box-shadow:0 1px 2px rgba(0,0,0,.04); }
.tile .num{ font-size:34px; font-weight:780; line-height:1; }
.tile .lab{ color:var(--muted); font-size:13px; margin-top:6px; text-transform:uppercase; letter-spacing:.04em; }
.tile.pass .num{ color:var(--pass);} .tile.warn .num{ color:var(--warn);} .tile.fail .num{ color:var(--fail);}
/* QC table -- overflow:visible so the custom hover tooltip is not clipped;
   corners are rounded on the header cells directly instead. */
table.qc{ width:100%; border-collapse:separate; border-spacing:0; font-size:14px;
  border-radius:12px; border:1px solid var(--line); }
table.qc th{ background:var(--ink); color:var(--card); text-align:left; padding:10px 12px; font-weight:650; }
table.qc th:first-child{ border-top-left-radius:11px; } table.qc th:last-child{ border-top-right-radius:11px; }
table.qc td{ padding:10px 12px; border-top:1px solid var(--line); vertical-align:top; }
table.qc tr:hover td{ filter:brightness(0.97); }
.badge{ display:inline-block; padding:3px 10px; border-radius:999px; font-weight:700; font-size:12px; letter-spacing:.03em; }
.badge.PASS{ background:var(--pass-bg); color:var(--pass);} .badge.WARN{ background:var(--warn-bg); color:var(--warn);} .badge.FAIL{ background:var(--fail-bg); color:var(--fail);}
tr.row-PASS td:first-child{ box-shadow:inset 4px 0 0 var(--pass);} tr.row-WARN td:first-child{ box-shadow:inset 4px 0 0 var(--warn);} tr.row-FAIL td:first-child{ box-shadow:inset 4px 0 0 var(--fail);}
/* Instant, styled hover tooltip on check names (native title[] is slow + clipped). */
.tip{ position:relative; border-bottom:1px dotted var(--muted); cursor:help; }
.tip:hover::after{
  content:attr(data-tip); position:absolute; left:0; top:150%; z-index:60;
  width:270px; white-space:normal; background:#1f2733; color:#fff;
  font-size:12.5px; font-weight:500; line-height:1.45; padding:9px 11px;
  border-radius:8px; box-shadow:0 8px 24px rgba(0,0,0,.30); pointer-events:none; }
.tip:hover::before{
  content:""; position:absolute; left:12px; top:135%; z-index:61;
  border:6px solid transparent; border-bottom-color:#1f2733; pointer-events:none; }
.forced{ background:var(--fail); color:#fff; padding:14px 18px; border-radius:12px; font-weight:700;
  box-shadow:0 8px 24px rgba(214,69,69,.35); margin:16px 0; }
.note{ color:var(--muted); font-size:14px; }
.legend{ font-size:13px; color:var(--muted); margin:8px 0 0; }
.legend span{ display:inline-block; margin-right:14px; }
.dot{ display:inline-block; width:10px; height:10px; border-radius:50%; margin-right:5px; vertical-align:middle; }
/* ---- DataTables (DT) widgets: make them follow the page theme ------------- */
/* Without this, DT ships its own light CSS and in dark mode the cells render as
   dark text on the dark page background (the rows look "blacked out"). */
.dataTables_wrapper{ color:var(--ink); margin:6px 0 10px; }
table.dataTable{ background:var(--card); border-collapse:separate!important; }
table.dataTable thead th{ color:var(--ink); border-bottom:2px solid var(--line); }
table.dataTable tbody tr{ background:var(--card); }
table.dataTable tbody td{ color:var(--ink); border-top:1px solid var(--line); }
table.dataTable.stripe tbody tr.odd, table.dataTable.display tbody tr.odd{ background:rgba(127,127,127,.06); }
table.dataTable tbody tr:hover, table.dataTable.hover tbody tr:hover{ background:rgba(79,70,229,.10); }
.dataTables_wrapper .dataTables_filter input,
.dataTables_wrapper .dataTables_length select{
  background:var(--card); color:var(--ink); border:1px solid var(--line);
  border-radius:6px; padding:3px 8px; }
.dataTables_wrapper .dataTables_info,
.dataTables_wrapper .dataTables_filter,
.dataTables_wrapper .dataTables_length,
.dataTables_wrapper .dataTables_paginate{ color:var(--muted)!important; }
.dataTables_wrapper .dataTables_paginate .paginate_button{ color:var(--ink)!important; }
.dataTables_wrapper .dataTables_paginate .paginate_button.current{
  background:var(--accent)!important; color:#fff!important; border-radius:6px; border:none; }
/* DT column-filter inputs (filter="top") */
table.dataTable thead .form-control, .dataTables_scrollHead input{
  background:var(--card); color:var(--ink); border:1px solid var(--line); }
```

```{r hero, results='asis'}
cat(sprintf('<div class="hero"><h1>RNA-seq Differential Expression Report</h1>
<div class="meta"><span>&#128197; <b>%s</b></span><span>&#9878;&#65039; Contrast: <b>%s</b></span>
<span>&#129516; Design: <b>%s</b></span><span>&#129440; %s</span></div></div>',
  params$run_timestamp, params$contrast,
  if (!is.null(cfg)) cfg$design else "", if (!is.null(cfg)) cfg$organism else ""))
```

```{r forced-banner, results='asis'}
if (isTRUE(params$forced))
  cat('<div class="forced">&#9888; THIS RUN WAS FORCED past one or more QC FAIL checks (--force). Interpret results with caution.</div>')
```

# QC panel

The traffic-light panel is the run's quality gate. **Green = PASS**, **amber =
WARN** (proceed, but review), **red = FAIL** (would stop the run without
`--force`). Hover a check name for what it means.

```{r qc-panel, results='asis'}
qc <- read_tsv_safe(p_qc)
# Short "why" tooltips for each known check.
tips <- c(
  sample_alignment      = "Do the count columns line up 1:1 with sample-sheet rows?",
  raw_count_sanity      = "Are values non-negative integers (raw counts, not TPM/CPM)?",
  library_size_floor    = "Hard floor on total reads per sample; below this a sample is unusable.",
  library_size_min      = "Recommended minimum total reads per sample.",
  genes_detected        = "Number of genes with >0 counts per sample.",
  replicates_per_group  = "Replicates per group in the tested factor; DE needs >=2.",
  design_identifiable   = "Is the model matrix full rank? Rank-deficient = a covariate is confounded.",
  low_count_filter      = "Fraction of genes removed by the low-count filter.",
  dispersion_fit        = "Sanity of the fitted dispersion trend.",
  cooks_outliers        = "Samples with many high Cook's-distance genes (candidate outliers)."
)
if (is.null(qc)) { cat("_QC results not found._") } else {
  n <- table(factor(qc$status, c("PASS","WARN","FAIL")))
  cat('<div class="tiles">')
  cat(sprintf('<div class="tile pass"><div class="num">%d</div><div class="lab">Pass</div></div>', n["PASS"]))
  cat(sprintf('<div class="tile warn"><div class="num">%d</div><div class="lab">Warn</div></div>', n["WARN"]))
  cat(sprintf('<div class="tile fail"><div class="num">%d</div><div class="lab">Fail</div></div>', n["FAIL"]))
  cat('</div>')
  cat('<table class="qc"><tr><th>Check</th><th>Status</th><th>Value</th><th>Details</th></tr>')
  for (i in seq_len(nrow(qc))) {
    s <- qc$status[i]; chk <- qc$check[i]
    tip <- if (!is.na(tips[chk])) sprintf(' data-tip="%s"', tips[chk]) else ""
    cls <- if (nzchar(tip)) "tip" else ""
    cat(sprintf('<tr class="row-%s"><td><span class="%s"%s>%s</span></td>', s, cls, tip, chk))
    cat(sprintf('<td><span class="badge %s">%s</span></td>', s, s))
    cat(sprintf('<td>%s</td><td>%s</td></tr>',
                ifelse(is.na(qc$value[i]),"",qc$value[i]), ifelse(is.na(qc$message[i]),"",qc$message[i])))
  }
  cat('</table>')
}
```

# Exploratory analysis

## PCA

PCA on the top-500 most variable variance-stabilized genes. Samples should
separate by biological condition; separation driven by batch is a red flag.
*Hover any point for its sample id and group.*

```{r pca}
# NOTE: htmlwidgets (plotly) must be RETURNED as the chunk value -- not cat()'d --
# so knitr's knit_print attaches the plotly.js dependency into the self-contained HTML.
sc <- read_tsv_safe(p_pca_tsv)
pv <- if (file.exists(p_pca_pv)) as.numeric(strsplit(readLines(p_pca_pv), "\t")[[1]]) else NULL
color_by <- if (!is.null(cfg)) cfg$contrast[[1]] else NULL
widget <- NULL
if (have_plotly && !is.null(sc) && all(c("PC1","PC2") %in% colnames(sc)) && !is.null(color_by) && color_by %in% colnames(sc)) {
  hover <- sprintf("<b>%s</b><br>%s: %s", sc$sample, color_by, sc[[color_by]])
  args <- list(data = sc, x = ~PC1, y = ~PC2, type = "scatter", mode = "markers",
               color = sc[[color_by]], text = hover, hoverinfo = "text",
               marker = list(size = 13, line = list(width = 1, color = "#ffffff")))
  bv <- if (nzchar(params$batch_vars)) trimws(strsplit(params$batch_vars, ",")[[1]])[1] else NA
  if (!is.na(bv) && bv %in% colnames(sc)) args$symbol <- sc[[bv]]
  widget <- plotly::layout(do.call(plotly::plot_ly, args),
    xaxis = list(title = if (!is.null(pv)) sprintf("PC1: %s%% variance", pv[1]) else "PC1", zeroline = FALSE),
    yaxis = list(title = if (!is.null(pv)) sprintf("PC2: %s%% variance", pv[2]) else "PC2", zeroline = FALSE),
    legend = list(title = list(text = color_by)))
}
if (!is.null(widget)) widget else if (file.exists(p_pca)) knitr::include_graphics(normalizePath(p_pca)) else knitr::asis_output("_PCA not available._")
```

```{r pca-batch, results='asis'}
if (file.exists(p_pca_bc)) { cat("**After batch correction (visualization only):**\n\n"); cat(img_tag(p_pca_bc, "PCA batch-corrected")) }
```

## Variance explained by metadata

Each cell is the R² of a principal component regressed on a metadata variable —
how much of that axis a variable explains. A technical variable dominating PC1
indicates a batch effect (handled inside the DE design). *Hover for exact R².*

```{r varmeta}
vm <- read_tsv_safe(p_varmeta, row.names = 1)
widget <- NULL
if (have_plotly && !is.null(vm) && ncol(vm) >= 1) {
  m <- as.matrix(vm)
  widget <- plotly::layout(
    plotly::plot_ly(x = colnames(m), y = rownames(m), z = m, type = "heatmap",
      colorscale = list(c(0,"#ffffff"), c(1,"#4f46e5")), zmin = 0, zmax = 1,
      text = matrix(sprintf("%s / %s<br>R^2 = %.2f",
              rep(rownames(m), ncol(m)), rep(colnames(m), each = nrow(m)), m), nrow = nrow(m)),
      hoverinfo = "text", colorbar = list(title = "R^2")),
    xaxis = list(title = ""), yaxis = list(title = "", autorange = "reversed"))
}
if (!is.null(widget)) widget else if (file.exists(p_vm_png)) knitr::include_graphics(normalizePath(p_vm_png)) else knitr::asis_output("_Not available._")
```

# Differential expression

```{r de-summary, results='asis'}
de <- read_tsv_safe(p_de)
if (is.null(de)) { cat("_DE table not found._") } else {
  sig <- de$significant %in% c(TRUE, "TRUE")
  n_sig <- sum(sig, na.rm = TRUE)
  n_up   <- sum(sig & de$log2FoldChange > 0, na.rm = TRUE)
  n_down <- sum(sig & de$log2FoldChange < 0, na.rm = TRUE)
  cat('<div class="tiles">')
  cat(sprintf('<div class="tile"><div class="num">%d</div><div class="lab">Genes tested</div></div>', nrow(de)))
  cat(sprintf('<div class="tile"><div class="num" style="color:%s">%d</div><div class="lab">Significant</div></div>', PAL$accent, n_sig))
  cat(sprintf('<div class="tile"><div class="num" style="color:%s">%d</div><div class="lab">Up in %s</div></div>', PAL$up, n_up, cfg$contrast[[2]]))
  cat(sprintf('<div class="tile"><div class="num" style="color:%s">%d</div><div class="lab">Down</div></div>', PAL$down, n_down))
  cat('</div>')
  cat(sprintf('<p class="note">Significance = padj &lt; %.3g <b>and</b> |log2FC| &ge; %.2g. Columns: <b>baseMean</b> mean normalized count; <b>log2FoldChange</b> shrunken log2(%s/%s); <b>lfcSE</b> its standard error; <b>padj</b> Benjamini-Hochberg adjusted p (use this for significance).</p>',
      cfg$de$padj_cutoff, cfg$de$lfc_cutoff, cfg$contrast[[2]], cfg$contrast[[3]]))
}
```

## Volcano

*Interactive — hover any point for gene, fold-change and adjusted p; drag to zoom.*

```{r volcano}
widget <- NULL
if (have_plotly && !is.null(de)) {
  d <- de[!is.na(de$padj), ]
  if (nrow(d)) {
    d$neglog10 <- -log10(pmax(d$padj, .Machine$double.xmin))
    sig <- d$significant %in% c(TRUE, "TRUE")
    d$dir <- ifelse(sig & d$log2FoldChange > 0, "Up",
              ifelse(sig & d$log2FoldChange < 0, "Down", "n.s."))
    d$lab <- if ("symbol" %in% colnames(d)) ifelse(is.na(d$symbol) | d$symbol=="", d$gene_id, d$symbol) else d$gene_id
    d$hover <- sprintf("<b>%s</b><br>log2FC: %.2f<br>padj: %.2g", d$lab, d$log2FoldChange, d$padj)
    lfc <- cfg$de$lfc_cutoff; pcut <- -log10(cfg$de$padj_cutoff)
    line <- function(x0,x1,y0,y1) list(type="line", x0=x0,x1=x1,y0=y0,y1=y1,
        line=list(dash="dash", color="#9aa6b6", width=1))
    rng <- range(d$log2FoldChange, na.rm = TRUE)
    widget <- plotly::layout(
      plotly::plot_ly(d, x = ~log2FoldChange, y = ~neglog10, color = ~dir,
        colors = c(Up = PAL$up, Down = PAL$down, "n.s." = PAL$ns),
        text = ~hover, hoverinfo = "text", type = "scatter", mode = "markers",
        marker = list(size = 6, opacity = 0.75)),
      xaxis = list(title = "log2 fold change"), yaxis = list(title = "-log10 adjusted p"),
      shapes = list(line(-lfc,-lfc,0,max(d$neglog10)), line(lfc,lfc,0,max(d$neglog10)),
                    line(rng[1],rng[2],pcut,pcut)))
    # Label the top genes by significance so notable hits are visible at a glance
    # (every gene is still discoverable via hover and the searchable table below).
    topn <- head(d[order(d$padj), ], 12)
    widget <- plotly::add_annotations(widget, data = topn, x = ~log2FoldChange, y = ~neglog10,
        text = ~lab, showarrow = TRUE, arrowhead = 0, arrowwidth = 0.6, arrowcolor = "#9aa6b6",
        ax = 18, ay = -14, font = list(size = 10, color = "#5b6675"))
  }
}
if (!is.null(widget)) widget else if (file.exists(p_volcano)) knitr::include_graphics(normalizePath(p_volcano)) else knitr::asis_output("_Not available._")
```

## Top genes

```{r de-table}
if (!is.null(de)) {
  cols <- intersect(c("gene_id","symbol","baseMean","log2FoldChange","lfcSE","pvalue","padj","significant"), colnames(de))
  d <- de[, cols]; numc <- vapply(d, is.numeric, logical(1))
  d[numc] <- lapply(d[numc], function(x) signif(x, 4))
  # Normalise the significance flag to the literal strings "TRUE"/"FALSE" so the
  # styleEqual() row highlight matches regardless of how read.delim typed it.
  if ("significant" %in% cols) d$significant <- ifelse(d$significant %in% c(TRUE, "TRUE"), "TRUE", "FALSE")
  if (have_DT) {
    dt <- DT::datatable(d, rownames = FALSE, filter = "top", extensions = "Buttons",
      class = "stripe hover row-border",
      options = list(pageLength = 15, order = list(list(which(cols=="padj")-1, "asc")),
                     dom = "Bfrtip", buttons = c("copy","csv"), scrollX = TRUE),
      caption = "Full DE results - search, sort, and export.")
    if ("significant" %in% cols)
      dt <- DT::formatStyle(dt, "significant", target = "row",
             backgroundColor = DT::styleEqual("TRUE", "rgba(79,70,229,.16)", default = ""),
             fontWeight = DT::styleEqual("TRUE", "600", default = "normal"))
    # Clean scientific display (3 sig figs) so tiny padj values don't show
    # floating-point artifacts like 7.3199999e-132.
    numcols <- intersect(c("baseMean","log2FoldChange","lfcSE","pvalue","padj"), cols)
    dt <- DT::formatSignif(dt, columns = numcols, digits = 3)
    htmltools::tagList(dt)
  } else knitr::kable(head(d, 25), row.names = FALSE)
}
```

# Functional enrichment

```{r enrich-dot, results='asis'}
ora <- read_tsv_safe(p_ora); gsea <- read_tsv_safe(p_gsea)
if (is.null(ora) && is.null(gsea))
  cat('<p class="note">Enrichment was disabled, produced no significant terms, or was skipped (see the run log).</p>')
if (file.exists(p_dot)) cat(img_tag(p_dot, "enrichment dotplot"))
```

```{r ora-table}
if (!is.null(ora) && nrow(ora) > 0) {
  cols <- intersect(c("ID","Description","GeneRatio","p.adjust","Count"), colnames(ora))
  d <- ora[, cols]; if ("p.adjust" %in% cols) d$p.adjust <- signif(d$p.adjust, 3)
  if (have_DT) htmltools::tagList(DT::datatable(d, rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE), caption = "ORA - over-represented GO terms"))
  else knitr::kable(head(d, 15), row.names = FALSE)
}
```

```{r gsea-table}
if (!is.null(gsea) && nrow(gsea) > 0) {
  cols <- intersect(c("ID","Description","setSize","NES","p.adjust"), colnames(gsea))
  d <- gsea[, cols]; numc <- vapply(d, is.numeric, logical(1)); d[numc] <- lapply(d[numc], signif, 3)
  if (have_DT) htmltools::tagList(DT::datatable(d, rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE), caption = "GSEA - enriched gene sets"))
  else knitr::kable(head(d, 15), row.names = FALSE)
}
```

# Reproducibility

```{r repro, results='asis'}
cat('<div class="card">')
cat(sprintf("<b>Seed:</b> %s<br>", if (!is.null(cfg)) cfg$seed else ""))
cat(sprintf("<b>Run timestamp:</b> %s<br>", params$run_timestamp))
cat(sprintf("<b>Organism / OrgDb:</b> %s<br>", if (!is.null(cfg)) cfg$organism else ""))
cat(sprintf("<b>Batch variable(s):</b> %s<br>", ifelse(nzchar(params$batch_vars), params$batch_vars, "none")))
cat(sprintf("<b>Full log + sessionInfo():</b> <code>%s</code><br>", file.path(outdir, "logs")))
cat("<b>Exact package versions:</b> see <code>renv.lock</code> in the project root.")
cat('</div>')
```
